Distortion corrected kernel density estimate on Riemannian manifolds

### 2022-11-07 paper/paper.Rmd
section 4 tidied up

Unlike the simulations where the ground truth densities are known, in the applications, we usually have little knowledge about the anomalous distributions of the households. Therefore, we assume, in this case, the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE. The results are summarized in \autoref{tab:electricityhdrpercent} similar to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) in $>99\%$ and $99\%$ HDRs that are also identified in different dimension reduction methods. Both DC-KDE and KDE are shown in the table. Except the lower percentage of ISOMAP in $>99\%$ HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods. The improvement brought by distortion correction is much more obvious in Laplacian Eigenmaps and t-SNE when the increase in both HDRs are greater than $0.5$.
\autoref{fig:electricityrank} presents the visualization of density ranking in log scale. The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.

@@

Now let us take a close look at the typical and anomalous distributions of households. According to \autoref{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration. Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}. Compared with ID 2396, the other three households show different usage patterns. ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh. As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household. The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days. It is likely that these two households are not occupied all the time during the trial and could be used as a AirBnb place or holiday place.

@@

In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities. Compared with KDE, our distortion corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding. The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point. We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension redution algorithm.

>>>>

Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households. To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE. The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) where anomalies fall outside the $>99\%$ and $99\%$ HDRs respectively that are also identified as anomalies by different dimension reduction methods. Both DC-KDE and KDE are shown in the table. Apart from the lower percentage of ISOMAP in $>99\%$ HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods. The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost comletely different anomalies to UMAP. \autoref{fig:electricityrank} presents the visualization of density ranking in log scale. The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods. The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}. The household indices shown in darker shades of blue are identified as anomalies the most dimension reduction algorithms. It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.

@@

We now further examine households 3161, 4546, 7049, which appear as anomalies for at least three dimension reduction algorithms as well as  Households 2396, which is the most typical household with the highest density estimate. Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}. Compared with ID 2396, the other three households show different usage patterns. ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh. As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household. The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days. It is likely that these two households are not occupied all the time during the trial and could be used as temporary accomodation (e.g. through a holiday rental service such as AirBnB).

@@

In this paper, we propose a novel distortion correction method to estimate the density of an embedding from manifold learning algorithms and further identify outliers based on the densities. Compared with KDE, our distortion corrected KDE makes use of geometric information for each data point to correct the distortion induced by the embedding. The Riemannian metric is estimated with the Learn Metric algorithm to approximate the geodesic distance and volume density function locally at each point. We compare our proposed method with KDE by two simulation settings, a 2-D manifold embedded as a 3-D twin peaks shape and a 4-D manifold mapped in a 100-D ambient space, and show that DC-KDE could generate more accurate kernel density estimation is more robust to the choice of dimension reduction algorithm.

### 2022-11-07 paper/paper.tex
section 4 tidied up

Unlike the simulations where the ground truth densities are known, in the applications, we usually have little knowledge about the anomalous distributions of the households. Therefore, we assume, in this case, the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE. The results are summarized in \autoref{tab:electricityhdrpercent} similar to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) in \(>99\%\) and \(99\%\) HDRs that are also identified in different dimension reduction methods. Both DC-KDE and KDE are shown in the table. Except the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods. The improvement brought by distortion correction is much more obvious in Laplacian Eigenmaps and t-SNE when the increase in both HDRs are greater than \(0.5\).
\autoref{fig:electricityrank} presents the visualization of density ranking in log scale. The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods.

@@

Now let us take a close look at the typical and anomalous distributions of households. According to \autoref{tab:electricityoutliers}, we pick the three dark blue indexes that are identified as anomalies the most times, which are Household 3161, 4546, 7049, as well as the most typical household with the highest density estimate, Households 2396, for illustration. Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}. Compared with ID 2396, the other three households show different usage patterns. ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh. As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household. The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days. It is likely that these two households are not occupied all the time during the trial and could be used as a AirBnb place or holiday place.

>>>>

Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households. To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE. The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent} presents the proportions of `true' outliers (given by UMAP) where anomalies fall outside the \(>99\%\) and \(99\%\) HDRs respectively that are also identified as anomalies by different dimension reduction methods. Both DC-KDE and KDE are shown in the table. Apart from the lower percentage of ISOMAP in \(>99\%\) HDRs, distortion corrected KDE gives higher proportions of true outliers in all other methods. The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost comletely different anomalies to UMAP. \autoref{fig:electricityrank} presents the visualization of density ranking in log scale. The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods. The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}. The household indices shown in darker shades of blue are identified as anomalies the most dimension reduction algorithms. It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.

@@

We now further examine households 3161, 4546, 7049, which appear as anomalies for at least three dimension reduction algorithms as well as Households 2396, which is the most typical household with the highest density estimate. Their half-hourly electricity demands across the entire trial period are plotted in \autoref{fig:electricityraw}. Compared with ID 2396, the other three households show different usage patterns. ID 3161 has much less electricity usage in summers, mostly below 1kWh, than in winters, above 3kWh. As this data set does not include the energy used for cooling or heating systems, it is possible that there are other large electronic appliances used in winter for this household. The other two anomalous IDs, 4546 and 7049, however, have nearly zero kWh electricity consumption for many days and some spikes for other days. It is likely that these two households are not occupied all the time during the trial and could be used as temporary accomodation (e.g.~through a holiday rental service such as AirBnB).
